---
title: "Final Project"
author: Clayton Rawson
output: html_document
---

## [Home](https://crawson1987.github.io) | [About Me](https://crawson1987.github.io/About_Me) | [CV](https::/crawson1987.github.io/CV) | [Final Project](https://crawson1987.github.io/Final_Project/)

```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
``` 
  
```{r loading programs and equations, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(phyloseq)
library(readxl)
library(patchwork)
library(modelr)

plot_bar2 <- function (physeq, x = "Sample", y = "Abundance", fill = NULL,
                       title = NULL, facet_grid = NULL)
{
  mdf = psmelt(physeq)
  p = ggplot(mdf, aes_string(x = x, y = y, fill = fill))
  p = p + geom_bar(stat = "identity", position = "stack")
  p = p + theme(axis.text.x = element_text(angle = -90, hjust = 0))
  if (!is.null(facet_grid)) {
    p <- p + facet_grid(facet_grid)
  }
  if (!is.null(title)) {
    p <- p + ggtitle(title)
  }
  return(p)
}
```
```{r A, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
ps <- readRDS("Clayton_SRA-main/output/ps_object_not_cleaned.RDS")

meta <- read_xlsx("PRJNA664779.xlsx") 

sample.names(ps)
rank_names(ps)
```
# A Quatitative and Statistical Analysis of a Previous Published Paper on Parkinson's Disease and the Gut Mycobiome 
  
## Introduction
  
The goal of this research was to analyze the metadata from a recently published paper by using various statistical methods as well as charts and plots. The goal was to get an more complete view of the data and the interpretations that could be made from it. The research paper that I will be using is titled "The Gut Mycobiome in Parkinson's Disease." The paper was published in February of 2021 in the Journal of Parkinson's Disease (vol. 11, no.1, pp. 152-158) by Mihai S. Cirstea, Kristen Sundvick, Ella Golz, Adam C. Yu, Rozlyn C. T. Boutin, Daniel Kliger, Brett B. Finlay, and Silke Appel-Cresswell. 

## Data Analysis

The first step in my analysis look at the data and see what the data looked like. The following chart looked at the number of samples per OTU. 
```{r B, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
otu_table(ps) %>% rowSums() %>% plot()
```

From the chart above we can see that a lot of the samples are clustered at the bottom with not a lot taxonomic units with a high number of samples. There are a few that have much higher numbers, above 10000, but most of them are low with values much less than 5000.

Despite not being encouraged by the number of samples I continued on and decided to look at the data from some different angles to see how diverse the data was. Maybe it was more diverse then I thought it was despite the low numbers of samples. I started this by looking at the numbers of samples in each of the three groups: control, PD (disease state), and NA group that has samples that fell into no categories. I also decided to look at the data in terms phylum level diversity. 

```{r C, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
sam_data(ps)$disease_status
tax_table(ps)[,1] %>% table()

fung <- subset_taxa(ps, Kingdom  == "k__Fungi")
tax_table(fung)[,1] %>% table()

colSums(otu_table(fung)) %>% summary()

rowSums(otu_table(fung)) %>% summary()

data.frame(SamplingEffort = rowSums(otu_table(fung)),
           Disease_Status = fung@sam_data$disease_status) %>% 
  ggplot(aes(x=factor(Disease_Status), y = SamplingEffort)) +
  geom_boxplot() + geom_jitter()

fung <- subset_samples(fung, !is.na(disease_status))

fung_ra <- transform_sample_counts(fung, function(x){x/sum(x)})
```
```{r E, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
plot_bar(fung_ra,fill = "Phylum")
```
```{r F, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
fung %>% 
  merge_samples("disease_status") %>% 
  transform_sample_counts(function(x){x/sum(x)}) %>% 
  plot_bar2(fill = "Phylum") + 
  theme_minimal() + 
  labs(y = "Relative Abundance")
``` 

The graph and chart above both helped me learn more about the metadata that the paper was working with. In terms of the chart, I found what I was expecting. The data is not very diverse in both the control and the PD groups.
The phylum level data looked reasonable. A saw the basic phyla that I was expecting to see. All of the main fungal phyla were present in numbers that I could see as reasonable. 

### Class Level Data

I decided that I wanted to next see the class level data and see how it compared to the diveristy I was seeing in the phylum level. 

```{r G, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
fung %>% 
  merge_samples("disease_status") %>% 
  transform_sample_counts(function(x){x/sum(x)}) %>% 
  plot_bar2(fill = "Class") + 
  theme_minimal() + 
  labs(y = "Relative Abundance") +
  scale_fill_viridis_d()
```

text place holder (not finished)

### Alpha Diversity

text place holder (not finished)

```{r H, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
fung %>% 
  phyloseq::plot_richness(x="disease_status", measures = c("Observed","Shannon")) +
  geom_boxplot()
```

text place holder (not finished) 

### Beta Diversity

text place holder (not finished)

```{r I, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
DCA <- fung %>% 
  transform_sample_counts(function(x){x/sum(x)}) %>% 
  ordinate(method = "PCoA")

plot_ordination(fung %>% transform_sample_counts(function(x){x/sum(x)}), 
                DCA,
                color="disease_status") +
theme_minimal()
```

text place holder (not finished)

### Categorical Comparison of BMI, sex, and disease status

text place holder (not finished)

```{r J, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
plotord2 <- function(i){
  p <- plot_ordination(fung %>% transform_sample_counts(function(x){x/sum(x)}), 
                       DCA,
                       color=i) +
    theme_minimal()
  return(p)
  
}

```{r K, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
p1 <- plotord2("BMI")
p2 <- plotord2("sex")
p3 <- plotord2("disease_status")

p1 + p2 + p3
```

```{r L, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
fung@sam_data$disease_status[fung@sam_data$disease_status == "PD"] <- 1
fung@sam_data$disease_status[fung@sam_data$disease_status == "Control"] <- 0
fung@sam_data$disease_status <- as.numeric(fung@sam_data$disease_statu)
```

### Shannon Diversity

text place holder (not finished)

```{r M, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
test <- estimate_richness(fung, measures = "Shannon")
test$Shannon
fung@sam_data$Shannon <- test$Shannon

mod <- glm(data=fung@sam_data %>% as("data.frame"),
           formula = disease_status ~ Shannon,
           family =  "binomial") 
```

text place holder (not finished)

```{r N, echo=FALSE, message=FALSE, warning=FALSE, include=TRUE}
add_predictions(data=fung@sam_data %>% as("data.frame"),
                model = mod, 
                type = "response") %>% 
  ggplot(aes(x=Shannon, y=pred,color=sex))+
  geom_point()
summary(mod)
```
  
text place holder (not finished)
  
# Conclusions 
  
text place holder (not finished)